---
title: "20074027 STAT2003 TASK 2"
author: "Bruce Odek Omondi"
date: "2025-05-12"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# QUESTION B
## 1. Part 1: Analysis for 16-run design only with a 2-level factorial design

Opening the dataset
```{r}
# Load required packages
library(tidyverse)

# Read data
drugdata <- read.csv("antiviraldrugs.csv")
```


Start by subsetting the dataset to include only the first 16 runs for 2-level factorial design
```{r}
# Subset only the first 16 runs
drugdata_16 <- drugdata[1:16, ]
```

Then define the model that will be used to get the effect estimates
```{r}
# get the average readout and encode replicates as -1 and 1 
drugs16 <- drugdata_16 %>%
  mutate(avg_readout = (Replicate1 + Replicate2) / 2,
         replicate = rep(c(-1, 1), each = 8))

# model on the first 16 runs
model16 <- lm(avg_readout ~ A + B + C + D + E +
                  A:B + A:C + A:D + A:E +
                  B:C + B:D + B:E +
                  C:D + C:E + D:E +
                  replicate,
                data = drugs16)

summary(model16)
```

The model turns out to be saturated (no degrees of freedom, p-val, etc.) so we cannot get the significance of the estimates. According to the textbook, this situation calls for the use of fold-over for the factorial design.

Fixing the 16-run design using fold-over to allow us to get the significance of the effect estimates
```{r}
# load libraries
library(tidyverse)

# getting only the first 16 runs 
original16 <- drugdata[1:16, ] %>%
  mutate(
    avg_readout = (Replicate1 + Replicate2) / 2,  # average of Rep 1 and 2
    replicate = rep(c(-1, 1), each = 8)           # encode replicates as -1 and 1
  )
```


Fold-over requires us to flip the signs of all the levels for each factor 
```{r}
# create folded-over design by reversing the signs of all factor levels
foldover16 <- original16 %>%
  mutate(
    A = -A, B = -B, C = -C, D = -D, E = -E,        # reverse signs as required for fold-over
    Run = Run + 100                                # label the foldover to differentiate from first 16
  )
```


Since we don't have actual data for the fold-over runs, I opted to simulate readout values for the fold-over runs. Ideally, these would be actual data from doing the fold-over during the experiment
```{r}
# simulate placeholder readout values for fold-over runs
set.seed(123)  # for reproducibility

foldover16$avg_readout <- rnorm(16, mean = mean(original16$avg_readout), sd = sd(original16$avg_readout))
foldover16$replicate <- rep(c(-1, 1), each = 8)

# combine original and fold-over data into one dataset which will have 32 runs total
folded_data <- bind_rows(original16, foldover16)

# fit full second-order interaction model using the combined data 
model_folded <- lm(avg_readout ~ A + B + C + D + E +
                                 A:B + A:C + A:D + A:E +
                                 B:C + B:D + B:E +
                                 C:D + C:E + D:E +
                                 replicate,
                   data = folded_data)
```


Then I defined a function that would be used to plot the half-normal plots of effects
```{r}
# define and generate a half-normal plot to assess effect significance
custom_half_normal <- function(model, model_name) {
  effects <- coef(model)[-1]  # exclude intercept
  abs_effects <- sort(abs(effects), decreasing = FALSE)
  n <- length(abs_effects)
  quantiles <- qnorm((1:n - 0.5) / n)
  
  plot(quantiles, abs_effects,
       main = paste("Half-Normal Plot –", model_name),
       xlab = "Normal Quantiles", ylab = "Absolute Effect", pch = 19)
  text(quantiles, abs_effects, labels = names(abs_effects), pos = 3, cex = 0.7)
  abline(lm(abs_effects ~ quantiles), col = "blue", lty = 2)
}

# output the summary and the plots
summary(model_folded)                    # Check model estimates
custom_half_normal(model_folded, "Foldover 32-Run Model")  # Plot effect significance

```

Running some model diagnostic checks
```{r}
# Plot standard diagnostic plots starting with the basic 4 plots
par(mfrow = c(2, 2))
plot(model_folded)

# and some additional checks for normality
# Histogram of residuals
par(mfrow = c(1, 1))  # Reset layout
hist(residuals(model_folded), main = "Histogram of Residuals",
     xlab = "Residuals", col = "lightblue", breaks = 10)

# Shapiro-Wilk test
shapiro.test(residuals(model_folded))
```



## 2. Part 2(Bonus marks): Analysis to replicate estimates a,b, and c in Table III using all 34 runs (16-run and 18-run designs combined as they did in the paper)

I. Estimates in Table III (a) with y=readout and all runs included
```{r}
# we use the average readout, so y= readout
drugs_a <- drugdata %>%
  mutate(avg_readout = (Replicate1 + Replicate2) / 2,
         replicate = rep(c(-1, 1), each = 17))

# Fit full second-order model using raw response
model_a <- lm(avg_readout ~ A + B + C + D + E +
                I(A^2) + I(B^2) + I(C^2) + I(D^2) + I(E^2) +
                A:B + A:C + A:D + A:E +
                B:C + B:D + B:E +
                C:D + C:E + D:E +
                replicate,
              data = drugs_a)

# Summary of model (estimates should match Table III(a))
summary(model_a)

```


II. Estimates in Table III (b) with y= sqrt(readout) after transformation and all runs included 
```{r}
# use the data but with y= sqrt(readout)
drugs_b <- drugdata %>%
  mutate(avg_readout = (Replicate1 + Replicate2) / 2,
         sqrt_readout = sqrt(avg_readout),
         replicate = rep(c(-1, 1), each = 17))

# fit model using sqrt(readout)
model_b <- lm(sqrt_readout ~ A + B + C + D + E +
                I(A^2) + I(B^2) + I(C^2) + I(D^2) + I(E^2) +
                A:B + A:C + A:D + A:E +
                B:C + B:D + B:E +
                C:D + C:E + D:E +
                replicate,
              data = drugs_b)

# Summary of model (estimates should match Table III(b))
summary(model_b)
```


III. Estimates in Table III (c) with y= sqrt(readout) and Replicate 1 of run 14 removed
```{r}
# Clean the dataset - remove replicate 1 of run 14 and use sqrt(readout)
drugs_c <- drugdata %>%
  mutate(Replicate1 = ifelse(Run == 14, NA, Replicate1)) %>%
  mutate(avg_readout = rowMeans(select(., Replicate1, Replicate2), na.rm = TRUE),
         sqrt_readout = sqrt(avg_readout),
         replicate = rep(c(-1, 1), each = 17))

# fit the full second-order model
model_c <- lm(sqrt_readout ~ A + B + C + D + E +
                I(A^2) + I(B^2) + I(C^2) + I(D^2) + I(E^2) +
                A:B + A:C + A:D + A:E +
                B:C + B:D + B:E +
                C:D + C:E + D:E +
                replicate,
              data = drugs_c)

# View model summary (same estimates as Table III(c))
summary(model_c)

# Diagnostic plots to check model assumptions
par(mfrow = c(2, 2))
plot(model_c)

```

Performing variable selection via stepwise regression to ensure there are no further significant effects before we re-plot the diagnostic plots to check for model assumptions.
```{r}
library(MASS)

# Stepwise regression to select significant effects
step_model <- stepAIC(model_c, direction = "both", trace = FALSE)

# final model output
summary(step_model)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(step_model)
```

Models a, b, and c outputs are mostly similar to the output in Table 3 of the study, as explained in the MS Word file. 



